---
title: "Activity Monitoring - Reproducible Research"
author: "Siddharth Gusain"
date: "October 1, 2020"
output:
  pdf_document: default
  html_document: default
---

##Loading and preprocessing the data

We will start by downloading the data from [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip), the data was downloaded at `r Sys.Date()`.

Methodology for downloading and preprocessing is given below.

```{r echo = TRUE, results = "hide", message = FALSE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",destfile = "activity.zip")
data <- read.csv(unz("activity.zip","activity.csv"),stringsAsFactors = FALSE)
library(lubridate)
library(dplyr)
library(ggplot2)
data$date<-ymd(data$date)
```

##Total number of steps
The summary of sum, mean and median of steps per day is given below.

```{r}
steps_summary_date <- data %>% group_by(date) %>% summarise(steps = sum (steps))
sumsteps <- sum(steps_summary_date$steps,na.rm = T)
meansteps <- as.integer(mean(steps_summary_date$steps,na.rm = T))
mediansteps <- median(steps_summary_date$steps,na.rm = T)
```
As calculated in the code chunk above, the sum is **`r sumsteps`**, the mean is **`r meansteps`** and the median is **`r mediansteps`**.

Furthermore, a histogram of the total number of steps per day is given below:
```{r echo = TRUE, message = FALSE}
plotdata <- data %>% group_by(date) %>% summarise(steps = sum(steps))
hist(plotdata$steps, breaks = 15, xlab = "Number of steps per day", main = "Histogram of daily steps per day", col = "green")
rug(plotdata$steps)
```

##Average daily activity pattern
A time series plot of 5-minute intervals against average number of steps taken across those intervals is given below.
```{r}
activitypatterndata <- data %>% select(steps,interval) %>% group_by(interval) %>% summarise(steps=mean(steps,na.rm = T))
plot(activitypatterndata$interval,activitypatterndata$steps,type="l",xlab = "5-minute intervals", ylab = "Average number of steps taken")
```

The details of interval wtih maximum number of average steps is given below.
```{r}
activitypatterndata[which.max(activitypatterndata$steps),]
```

##Imputing missing values
Using the sum function to calculate the total number of missing values in the "steps" variable as "date" and "interval"" have no missing values. 

```{r}
sumna <- sum(is.na(data$steps))
```
The total missing values in the steps variable of the dataset are **`r sumna`**.

I will now replace the "NA" values with the mean value of that particular interval's remaining values.

```{r}
replacewithmean <- function(x){replace(x, is.na(x), mean(x, na.rm = TRUE))}
completedata <- data %>% group_by(interval) %>% mutate( steps = replacewithmean(steps))
```
The total number of steps taken each day as per the dataset without missing values can be seen as below:

```{r echo = TRUE, message = FALSE}
plotdatacomplete <- completedata %>% group_by(date) %>% summarise(steps = sum(steps))
hist(plotdatacomplete$steps, breaks = 15, xlab = "Number of steps per day", main = "Histogram of daily steps per day", col = "blue")
rug(plotdatacomplete$steps)
```

```{r}
steps_summary_completedata <- completedata %>% group_by(date) %>% summarise(steps = sum (steps))
meantotal <- as.integer(mean(steps_summary_completedata$steps,na.rm = T))
mediantotal <- as.integer(median(steps_summary_completedata$steps,na.rm = T))
```
The mean of steps per day from the dataset without missing values is **`r meantotal`** and the median from the same dataset is **`r mediantotal`**.

```{r}
diffmean <- meantotal - meansteps
diffmedian <- mediantotal - mediansteps
```

The difference between means from the two data sets is `r `diffmean`and the difference between medians from the two data sets is `r `diffmedian`.

##Difference in activity patterns between weekdays and weekends

```{r}
completedata$day <- weekdays(completedata$date)
completedata$weekday <- ifelse(completedata$day == "Saturday" | completedata$day == "Sunday", "Weekend", "Weekday")
completedata$weekday <- as.factor(completedata$weekday)
completedata <- completedata %>% group_by(weekday,interval) %>% summarise(steps = mean(steps, na.rm=T))
qplot(x = interval, y = steps, data = completedata, facets = weekday~.,geom = "path",color = weekday)
```
