---
title: "Activity Monitoring - Reproducible Research"
author: "Siddharth Gusain"
date: "October 1, 2020"
output: html_document
---

##Loading and preprocessing the data

We will start by downloading the data from [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip), the data was downloaded at `r Sys.Date()`.

Methodology for downloading and preprocessing is given below.

```{r echo = TRUE, results = "hide", message = FALSE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",destfile = "activity.zip")
data <- read.csv(unz("activity.zip","activity.csv"),stringsAsFactors = FALSE)
library(lubridate)
library(dplyr)
data$date<-ymd(data$date)
```

##Total number of steps
The summary of sum, mean and median of steps per day is given below.

```{r}
steps_summary_date <- data %>% group_by(date) %>% summarise(steps = sum (steps))
sum(steps_summary_date$steps,na.rm = T)
mean(steps_summary_date$steps,na.rm = T)
median(steps_summary_date$steps,na.rm = T)
```

Furthermore, a histogram of the total number of steps per day is given below:
```{r echo = TRUE, message = FALSE}
plotdata <- data %>% group_by(date) %>% summarise(steps = sum(steps))
hist(plotdata$steps, breaks = 15, xlab = "Number of steps per day", main = "Histogram of daily steps per day", col = "green")
rug(plotdata$steps)
```

##Average daily activity pattern
A time series plot of 5-minute intervals against average number of steps taken across those intervals is given below.
```{r}
activitypatterndata <- data %>% select(steps,interval) %>% group_by(interval) %>% summarise(steps=mean(steps,na.rm = T))
plot(activitypatterndata$interval,activitypatterndata$steps,type="l",xlab = "5-minute intervals", ylab = "Average number of steps taken")
```

The details of interval wtih maximum number of average steps is given below.
```{r}
activitypatterndata[which.max(activitypatterndata$steps),]
```